---
title: "WORKSHOP_4_FULLDOC"
output: github_document
---

## Step 2: Assembly.

Started by making a new directory called flye using:
```{r, eval=FALSE}
mkdir flye
```


Change directory into flye and then make a symbolic link to the bbd read data:

```{r, eval=FALSE}
ln -s /pvol/data/metagenome_assembly/bbd_0.1.fastq.gz .
```


Shell script called run_flye_bbd.sh created inside inside flye directory: 
```{r, eval=FALSE}

#!/bin/bash
#SBATCH --time=720
#SBATCH --ntasks=12 --mem=10gb

echo "Starting flye bbd in $(pwd) at $(date)"

flye --nano-hq ../raw_data/bbd_0.1.fastq.gz --threads 12 --out-dir bbd --meta

echo "Finished flye bbd in $(pwd) at $(date)"
```

Submit batch job in terminal using:
```{r, eval=FALSE}
sbatch run_flye_bbd.sh

```


## Step 3 binning
Create a new metabat directory at the top level and change to the new directory:
```{r, eval=FALSE}
mkdir metabat
cd metabat
```


extract depth information and reorder contigs based on depth:
```{r, eval=FALSE}
cat ../flye/bbd/assembly_info.txt  | grep -v '#' | awk '{OFS="\t";print $1,$3}' > depths_bbd.txt # extract depth info

cat depths_bbd.txt | awk '{print $1}' | xargs samtools faidx ../flye/bbd/assembly.fasta > assembly_bbd_ordered.fasta # Reorder contigs based on depth

```

Run metabat using terminal command line:
```{r, eval=FALSE}
metabat2 -i assembly_bbd_ordered.fasta -a depths_bbd.txt --cvExt -o bins_bbd/metabat --seed 5

```


## Step 4: Check bin quality with checkm

Make a directory for checkm and cd into it:

```{r, eval=FALSE}
mkdir checkm
cd checkm
```

Find the number of the largest contig wihtin the bbd assembly and manually include within metabat, Command below displays the lengths of all contigs sorted from shortest to longest.
```{r, eval=FALSE}
cat ../flye/bbd/assembly.fasta | bioawk -c fastx '{print length($seq),$name}' | sort -n

```

largest contigs: 
4706701 contig_119

Substitute largest contig in the code below (contig_119):
```{r, eval=FALSE}
mkdir bins_bbd
cp ../metabat/bins_bbd/*.fa bins_bbd/
samtools faidx ../flye/bbd/assembly.fasta contig_119 > bins_bbd/contig_119.fa
```



Next create a slurm script for your checkm run and save within this directory as  run_checkm_bbd.sh
```{r, eval=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=6 --mem=40gb

echo "Starting checkm in $(pwd) at $(date)"

shopt -s expand_aliases
alias checkm='apptainer run -B /pvol/:/pvol /pvol/data/sif/checkm.sif checkm'

checkm lineage_wf bins_bbd out -t 4 -x fa -f checkm_results_bbd.txt

echo "Finished checkm in $(pwd) at $(date)"
```

Submit your checkm job like this

```{r, eval=FALSE}
sbatch run_checkm_bbd.sh

```


##Step 5 Select bins based on checkm results

Change directory to the checkm folder.
```{r, eval=FALSE}
cd checkm

```

checkm analysis. These are summarised in the file checkm_results_bbd.txt - keep in mind it is a text file so space count for a coloumn
```{r, eval=FALSE}
head checkm_results_bbd.txt 

```

Focusing on bins with a completeness greater than 15%. To extract the bin IDs for these we can use awk. This command should print only the lines where the 13th column has a value greater than 15.
```{r, eval=FALSE}
cat checkm_results_bbd.txt | awk '$13>15'

```

This seems to be the bins we want but we just want the bin ID not the entire line. To get this we can use the print command
```{r, eval=FALSE}
cat checkm_results_bbd.txt | awk '$13>15{print $1}'

```

Now let’s send this output to a file so that we can use it in the next step

```{r, eval=FALSE}
cat checkm_results_bbd.txt | awk '$13>15{print $1}' > bbd_topbins.txt

```


## Step 6: Perform taxonomic classification with gtdbtk
First create a folder for gtdbtk in the top level of your project. Then change directory to this folder.

```{r, eval=FALSE}
mkdir gtdbtk
cd gtdbtk
```

Creating a while loop for genomes 
```{r, eval=FALSE}
while read bin;do echo $bin;done < ../checkm/bbd_topbins.txt

while read bin;do echo "../metabat/bins_bbd/${bin}.fa";done < ../checkm/bbd_topbins.txt

while read bin;do cp "../metabat/bins_bbd/${bin}.fa" genomes;done < ../checkm/bbd_topbins.txt
```

Check that it worked by listing files within the genomes directory.
```{r, eval=FALSE}
ls genomes

```

First create a blank shell script file and save it inside your gtdbtk directory. Name your file run_gtdbtk.sh:

```{r, eval=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=40gb

echo "Starting gtdbtk in $(pwd) at $(date)"

shopt -s expand_aliases
alias gtdbtk='apptainer run -B /pvol:/pvol,/pvol/data/mg/gtdbtkdata/release207_v2/:/refdata /pvol/data/sif/gtdbtk.sif'

gtdbtk classify_wf -x fa --cpus 2 --genome_dir genomes --out_dir out --skip_ani_screen

echo "Finished gtdbtk in $(pwd) at $(date)"
```

Submit batch job:
```{r, eval=FALSE}
sbatch run_gtdbtk.sh

```


## Step 7. Find and annotate genes with prokka

Create new directory and change to it : 
```{r, eval=FALSE}
mkdir prokka

cd prokka 

```


Next find out the first bin ID in your list

```{r, eval=FALSE}
head -n 1 ../checkm/bbd_topbins.txt

```

first bin id is contig 119:

```{r, eval=FALSE}
bin=contig_119
prokka --outdir $bin --prefix $bin  ../metabat/bins_bbd/${bin}.fa
```


Using ls command to see files within the contig_119 files
```{r, eval=FALSE}
ls contig_119

```
- contig_119.err  contig_119.ffn  contig_119.fsa  contig_119.gff  contig_119.sqn  contig_119.tsv
contig_119.faa  contig_119.fna  contig_119.gbk  contig_119.log  contig_119.tbl  contig_119.txt

Create a new blank text file and save it in your prokka directory, called run_prokka.sh
```{r, eval=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=2gb

echo "Starting prokka in $(pwd) at $(date)"

shopt -s expand_aliases
alias prokka='apptainer run -B /pvol/:/pvol /pvol/data/sif/prokka.sif prokka'

while read bin;do 
  prokka --outdir $bin --prefix $bin "../metabat/bins_bbd/${bin}.fa";
done < ../checkm/bbd_topbins.txt

echo "Finished prokka in $(pwd) at $(date)"

```

submit batch job by:
```{r, eval=FALSE}
sbatch run_prokka.sh

```


## Step 8: Infer metabolic pathways

Create new directory called minpath and change to it:
```{r, eval=FALSE}
mkdir minpath
cd minpath
```

Inputs for minpath will be prepared for each bin based on the prokka .tsv file. Start by looking at the .tsv file for a single bin
inputs:
```{r, eval=FALSE}
head ../prokka/contig_119/contig_119.tsv

```

we only want to keep a unique ID for each gene (column 1) and its EC number (column 5). We can transform the tsv into this format using awk like this
```{r, eval=FALSE}
head ../prokka/contig_119/contig_119.tsv | awk -F '\t' '{print $1,$5}'

```

Note that here we are using -F '\t' to tell awk that our input is tab separated. We also want to filter the output to only include lines with an EC number
This command extracts gene IDs and their EC numbers from the Prokka annotation file, removes the header, and saves the tab-separated results for use in MinPath.
```{r, eval=FALSE}
head ../prokka/contig_119/contig_119.tsv | awk -F '\t' '$5{print $1,$5}'

head ../prokka/contig_119/contig_119.tsv | awk -F '\t' '$5{OFS="\t";print $1,$5}' | grep -v '^locus'

cat ../prokka/contig_119/contig_119.tsv | awk -F '\t' '$5{OFS="\t";print $1,$5}' | grep -v '^locus'

cat ../prokka/contig_119/contig_119.tsv | awk -F '\t' '$5{OFS="\t";print $1,$5}' | grep -v '^locus' > contig_119.ec
```


Now run minpath on this file as follows;

```{r, eval=FALSE}
minpath -ec contig_119.ec -report contig_119.ec.report -detail contig_119.ec.details

```


Take a look at the top few lines of output

```{r, eval=FALSE}
head contig_119.ec.report 

```

checking prokka with:
```{r, eval=FALSE}
tail ../prokka/slurm-607.out

```


Once prokka had finished running we run the following code to prepare minpath inputs for all bins

```{r, eval=FALSE}
while read bin;do 
  cat ../prokka/$bin/${bin}.tsv | awk -F '\t' '$5{printf("%s\t%s\n", $1,$5)}' | grep -v '^locus' > ${bin}.ec 
done < ../checkm/bbd_topbins.txt

while read bin;do 
  minpath -ec ${bin}.ec -report ${bin}.ec.report -details ${bin}.ec.details
done < ../checkm/bbd_topbins.txt
```


Use the following command to search for this pathway across all the bins

```{r, eval=FALSE}
grep 'PWY-101' *.ec.report

```

Question 4.1
How many bins contain at least one gene from this pathway?
-Every single contains at least one gene from this pathway

Which bin has all of the genes (ie a complete pathway)?
- Metabat.20 contains all 4 gene families in the pathway

contig_119.ec.report:path PWY-101 any n/a  naive 1  minpath 0  fam0  4  fam-found  1  name  photosynthesis light reactions
metabat.20.ec.report:path PWY-101 any n/a  naive 1  minpath 1  fam0  4  fam-found  4  name  photosynthesis light reactions
metabat.25.ec.report:path PWY-101 any n/a  naive 1  minpath 1  fam0  4  fam-found  1  name  photosynthesis light reactions
metabat.32.ec.report:path PWY-101 any n/a  naive 1  minpath 1  fam0  4  fam-found  1  name  photosynthesis light reactions
metabat.70.ec.report:path PWY-101 any n/a  naive 1  minpath 0  fam0  4  fam-found  1  name  photosynthesis light reactions

Cross reference your results with the output from gtdbtk to see the taxonomy of bins containing this pathway:
```{r, eval=FALSE}
grep 'metabat.20' ../gtdbtk/out/gtdbtk.bac120.summary.tsv

```

```{r, warning = F, message = F}
library(tidyverse)
gtdbtk_results <- read_tsv("gtdbtk/out/gtdbtk.bac120.summary.tsv") # create r object to view results more easily 
```

Question 4.2
How many bins contain the pathway PWY-5360?

The PWY-5360 contains only 1 bin.

What family (in a taxonomic sense) do these bins belong to?

family Arcobacteraceae, genus Halarcobacter

